Simple repression draft

### 2020-06-01 doc/section_05_discussion.tex
MRM pass through entire main text. Removed side notes. Still lots left

The present work has tackled that question in the context of the first
regulatory architecture hypothesized in the molecular biology era, namely, the
repressor-operator model of Jacob and Monod. Regulation in that architecture is
the result of a competition between a repressor which inhibits transcription and
RNAP polymerase which undertakes it. Through the labors of generations of
geneticists, molecular biologists and biochemists, an overwhelming amount of
information and insight has been garnered into this simple regulatory motif,
licensing it as what one might call the ``hydrogen atom'' of regulatory biology.
It is from that perspective that the present paper explores the extent to which
some of the different models that have been articulated to describe that motif
allow us to understand both the average level of gene expression found in a
population of cells, the intrinsic cell-to-cell variability, and the full gene
expression distribution found in such population as would be reported in a FISH
experiment, for example.

Our key insights can be summarized as follows. First, as shown in 
Figure~\ref{fig1:means_cartoons}, the mean expression in the simple repression
architecture is captured by a master curve in which the action of repressor and
the details of the polymerase interaction with the promoter appear separately
and additively in an effective free energy. Interestingly, as has been shown
elsewhere in the context of the Monod-Wyman-Changeux model, these kinds of
coarse-graining results are an exact mathematical result and do not constitute
hopeful approximations or biological naivete \cite{Razo-Mejia2018, Chure2019}.
To further dissect the relative merits of the different models, we must appeal
to higher moments of the gene expression probability distribution. To that end,
our second set of insights focus on gene expression noise, where it is seen that
a treatment of the constitutive promoter already reveals that some models have
Fano factors (variance/mean) that are less than one, at odds with
any and all experimental data that we are aware of.
This theoretical result allows us to directly discard a subset of the models
(models 1-3 in Figure~\ref{fig2:constit_cartoons}) since they cannot be
reconciled with experimental observations. The two remaining models (models 4 and 5
in Figure~\ref{fig2:constit_cartoons}) appear to contain enough microscopic
realistm to be able to reproduce the data. A previous exploration of one 
of these models demonstrated the sloppy nature of the model in which data on single-cell
mRNA counts alone cannot constrain the value of all parameters 
simultaneously \cite{Razo-Mejia2020}. Here we demonstrate that the proposed
one-state bursty promoter model (model 5 in 
Figure~\ref{fig2:constit_cartoons}) emerges as  a limit of the commonly used
two-state promoter model \cite{Peccoud1995}. We put  the idea to the test that this
level of coarse-graining is rich enough to reproduce previous 
experimental observations. In particular we perform Bayesian inference to determine the
two parameters describing the full steady-state mRNA distribution, finding that
the model is able to provide a quantitative description of  a plethora of promoter sequences with different
mean levels of expression and noise.

\mrm{
Here we should include a discussion of the regulated case.  RP: I made a first
brief stab at it.
}
With the results of the constitutive promoter in hand, we then fix the parameters
associated with this class of promoters and use them as input for evaluating
the noise in gene expression for the simple repression motif itself.  This analysis
allows us to provide a single overarching analysis of both the constitutive
and simple repression architectures using one simple model and
corresponding set of self-consistent parameters, demonstrating not only
a predictive framework, but also reconciling the equilibrium and non-equilibrium
views of the same simple repression constructs.

>>>>

The present work has tackled that question in the context of the first
regulatory architecture hypothesized in the molecular biology era, namely, the
repressor-operator model of Jacob and Monod. Regulation in that architecture is
the result of a competition between a repressor which inhibits transcription and
RNAP polymerase which undertakes it. Through the labors of generations of
geneticists, molecular biologists and biochemists, an overwhelming amount of
information and insight has been garnered into this simple regulatory motif,
licensing it as what one might call the ``hydrogen atom'' of regulatory biology.
It is from that perspective that the present paper explores the extent to which
some of the different models that have been articulated to describe that motif
allow us to understand both the average level of gene expression found in a
population of cells, the intrinsic cell-to-cell variability, and the full gene
expression distribution found in such population as would be reported in a
single molecule Fluorescence in situ Hybridization experiment, for example.

Our key insights can be summarized as follows. First, as shown in
Figure~\ref{fig1:means_cartoons}, the mean expression in the simple repression
architecture is captured by a master curve in which the action of repressor and
the details of the polymerase interaction with the promoter appear separately
and additively in an effective free energy. Interestingly, as has been shown
elsewhere in the context of the Monod-Wyman-Changeux model, these kinds of
coarse-graining results are an exact mathematical result and do not constitute
hopeful approximations or biological naivete \cite{Razo-Mejia2018, Chure2019}.
To further dissect the relative merits of the different models, we must appeal
to higher moments of the gene expression probability distribution. To that end,
our second set of insights focus on gene expression noise, where it is seen that
a treatment of the constitutive promoter already reveals that some models have
Fano factors (variance/mean) that are less than one, at odds with any and all
experimental data that we are aware of. This theoretical result allows us to
directly discard a subset of the models (models 1-3 in
Figure~\ref{fig2:constit_cartoons}(A)) since they cannot be reconciled with
experimental observations. The two remaining models (models 4 and 5 in
Figure~\ref{fig2:constit_cartoons}) appear to contain enough microscopic
realistm to be able to reproduce the data. A previous exploration of one of
these models demonstrated the sloppy nature of the model in which data on
single-cell mRNA counts alone cannot constrain the value of all parameters
simultaneously \cite{Razo-Mejia2020}. Here we demonstrate that the proposed
one-state bursty promoter model (model 5 in Figure~\ref{fig2:constit_cartoons})
emerges as  a limit of the commonly used two-state promoter model
\cite{Peccoud1995}. We put  the idea to the test that this level of
coarse-graining is rich enough to reproduce previous experimental observations.
In particular we perform Bayesian inference to determine the two parameters
describing the full steady-state mRNA distribution, finding that the model is
able to provide a quantitative description of  a plethora of promoter sequences
with different mean levels of expression and noise.

With the results of the constitutive promoter in hand, we then fix the
parameters associated with this class of promoters and use them as input for
evaluating the noise in gene expression for the simple repression motif itself.
This analysis allows us to provide a single overarching analysis of both the
constitutive and simple repression architectures using one simple model and
corresponding set of self-consistent parameters, demonstrating not only a
predictive framework, but also reconciling the equilibrium and non-equilibrium
views of the same simple repression constructs.
